---
output: 
  github_document
---

<!-- Edit the README.Rmd only!!! The README.md is generated automatically from README.Rmd. -->

kehra: an R package to collect, assemble and model air pollution, weather and health data
==============================================================

```{r echo=FALSE}
knitr::opts_chunk$set(
  comment = '#>',
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  eval = FALSE,
  cache = TRUE
)
```

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.154722.svg)](https://doi.org/10.5281/zenodo.154722)

[![Travis-CI Build Status](https://travis-ci.org/kehraProject/kehra.svg?branch=master)](https://travis-ci.org/kehraProject/kehra)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/kehraProject/kehra?branch=master&svg=true)](https://ci.appveyor.com/project/kehraProject/kehra)
[![Coverage Status](https://img.shields.io/codecov/c/github/kehraProject/kehra/master.svg)](https://codecov.io/github/kehraProject/kehra?branch=master)

[![CRAN Status Badge](http://www.r-pkg.org/badges/version/kehra)](http://cran.r-project.org/web/packages/kehra)
[![CRAN Total Downloads](http://cranlogs.r-pkg.org/badges/grand-total/kehra)](http://cran.rstudio.com/web/packages/kehra/index.html)
[![CRAN Monthly Downloads](http://cranlogs.r-pkg.org/badges/kehra)](http://cran.rstudio.com/web/packages/kehra/index.html)

The package [kehra](https://cran.r-project.org/package=kehra) allows to collect, assemble and model air pollution, weather and health data. The package is being developed as part of the British Council's funded KEHRA project and builds upon other R packages such as [rdefra](https://cran.r-project.org/package=rdefra) to retrieve pollution data from the UK-AIR database hosted by the UK Department for Enviornment, Food and Rural Affairs, and the [bnlearn](https://cran.r-project.org/package=bnlearn) to model the assembled database using Bayesian Networks.

## Dependencies

The kehra package depends on the Geospatial Data Abstraction Library (gdal) and some additional CRAN packages. Check for missing dependencies and install them using the commands below:

```{r}
packs <- c('Hmisc', 'raster', 'reshape2', 'stringr', 'sp', 'xts', 'zoo', 
           'devtools', 'rgdal')
new.packages <- packs[!(packs %in% installed.packages()[,'Package'])]
if(length(new.packages)) install.packages(new.packages)
```

## Installation

Get the released version from CRAN:

```{r}
install.packages('kehra')
```

Or the development version from github using [devtools](https://github.com/hadley/devtools):

```{r}
devtools::install_github('kehraProject/kehra')
```

Load the kehra package:

```{r, eval = TRUE}
library('kehra')
```

## Meta

* Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.
* Please [report any issues or bugs](https://github.com/kehraProject/kehra/issues).
* License: [GPL-3](https://opensource.org/licenses/GPL-3.0)
* Get citation information for `kehra` in R doing `citation(package = 'kehra')`
